Changes in the justification of educational inequalities. The role of perceptions of inequality and meritocracy during the COVID pandemic.

### 2024-12-20 01-introduction.qmd
cambios intro+antecedentes

Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated. As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023]. Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021]. Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023]. In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain.

Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background. In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022]. This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008]. Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile, and to what extent this phenomenon is associated with the perception of meritocracy and its variations during the COVID pandemic.

@@

This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework. Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic [@hankivsky_introduction_2022;@breznau_welfare_2021]. A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises. Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals. Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they are the most affected in critical times.

>>>>

Despite the widespread recognition of education as a key factor for social mobility and equality of opportunities, disparities in this domain are far from being eliminated. As the last PISA Report points out, “Socio-economically disadvantaged students are seven times more likely than advantaged students to score below Level 2 in mathematics on average across OECD countries” [@oecd_pisa_2023]. Furthermore, empirical evidence from social science research has revealed an exacerbation of educational inequalities across various domains, such as family socioeconomic status, gender, ethnicity, and migratory background, particularly after COVID-19 [@darmodyImpactsCOVID19Control2021]. Nevertheless, and despite the consistent empirical evidence on educational inequalities and their consequences, research on distributive justice indicates remarkable individual differences in the justification of unequal access to education [@leeFairnessPerceptionsEducational2023]. In the present paper, we argue that the perception of meritocracy in society might play a role in the justification of economic inequalities in this particular domain. We examine this association in the context of COVID 19 in Chile, a period characterized by several challenges regarding the political and economic system that could have affected the way in which the role of individual performance and social outcomes are deemed.

Meritocracy is defined as a distributive system where individual factors such as effort and talent are conceived as the main determinants of individual performance, usually sidelining the role of opportunities and social background. In recent years, there has been a growing interest in studying meritocratic beliefs as well as their variations according to individual and contextual variables [@mijs_paradox_2021; @mijs_belief_2022]. This line of research has highlighted how meritocratic beliefs are associated with justifying unequal access to resources and rewards in society, by legitimizing social hierarchies in which those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004;@hadjar_meritokratie_2008]. Within this framework, this study focuses on the changes in the justification of inequalities in access to education in Chile during the COVID pandemic, and to what extent this phenomenon is associated with the perception of meritocracy and its variations.

@@

This research endeavors to dissect the nuanced alterations in the justifications for educational disparities within the Chilean framework. Central to our investigation is the premise that the rationale underpinning these inequalities experiences a significant recalibration during periods of vulnerability and crises, specifically under the unprecedented strains of the global health and economic cataclysm triggered by the COVID-19 pandemic [@hankivsky_introduction_2022;@breznau_welfare_2021]. A pertinent aspect of this discourse is articulated in the works of Mijs [-@mijs_stratified_2016;-@mijs_visualizing_2018], showing the interplay between meritocratic beliefs and the legitimization of disparities, as well as the potential erosion of meritocratic principles amidst crises. Within this framework, we point out that periods of vulnerability (such as the health and economic crisis resulting from the COVID-19 pandemic) could weaken the justification of social inequalities, a change that would be at least in part linked to a challenge of the meritocratic ideals. Furthermore, we suggest that the association between meritocratic perception and inequality justification would be weaker among lower-status groups, as they would be the most affected in critical times.

[[no sé si de pronto aclarar más la idea de que no es posible aislar el efecto COVID propiamente tal ...]]

### 2024-12-20 02-antecedentes.qmd
cambios intro+antecedentes

# Inequality, beliefs and justification

@@

The field of studies of subjective economic inequality has been extensively developed in the social sciences, particularly in the area of social justice research [@Resh2016]. From a sociological point of view, the works of @kluegelBeliefsStratification1981 can be considered pioneers in framing a research area comprising perceptions, beliefs, and justifications in topics such as inequality of opportunities, poverty, and distributive justice. Theoretically, the authors understand the concept of "beliefs" in a "broad sense to refer to the information (veridical or non-veridical) about a phenomenon that an individual uses as a basis both for inferring other information and for action." [@kluegelBeliefsStratification1981, p. 30]. However, the literature has strained the concept of "beliefs" by more specific distinctions around the individual and structural dimensions of economic inequality. In this regard, @wegenerIllusionDistributiveJustice1987 has suggested that perceptions of social status and reward hierarchy are relative to the observer's status in such hierarchies [@wegenerIllusionDistributiveJustice1987, p. 2]. This approach established that status positions influence how people perceive things _are_ and, in turn, these perceptions anchor how people evaluate this inequality as just or unjust [@jassoJusticeEarningsNew1978]. Additionally, @janmaatSubjectiveInequalityReview2013 has suggested more detailed conceptual distinctions about how people perceive economic inequality _is_, how they believe or prefer it _should be_, and the process of evaluating such differences as fair.

 Theoretically, @liebigSociologyJustice2016 establish three fundamental distinctions in justice attitudes in the field of sociological justice research. First, _order-related attitudes_ refer to individual motives that explain preferences for resource allocation based on the principles of equality, need, or entitlement. Second, _procedural justice attitudes_ account for the degree of acceptance of unfavorable outcomes under conditions in which the procedure under which those decisions were made is considered fair (e.g., opportunities). Finally, they can be _outcome-related attitudes_ which refer primarily to the economic resources that a person currently obtains and the set of collective commitments associated with possessing those resources (e.g., transfers or taxes). Within this framework, we conceptualize the justification of inequality in the first area of order-related attitudes, as it expresses preferences for resource allocation that can be associated with concepts that reflect different distributive principles [@kluegel_social_1995;@sonhing_merit_2011]

An important finding in the area of order-related attitudes is that beliefs about inequality are stratified by social status [@janmaatSubjectiveInequalityReview2013; @kluegelBeliefsInequalityAmericans1986]. The evidence suggests that advantaged positions in terms of socioeconomic status, such as income, education, or occupational prestige, are associated with preferences for less egalitarian class imageries [@Evans1992;@Evans2017] and greater justification of salary gaps [@hadler_why_2005;@kelley_legitimate_2021]. Nevertheless, and in contrast to this perspective, it has been argued that human beings are intrinsically motivated to justify the system they inhabit, even when the information available to them points in the opposite direction [@jost_psychology_2003;@vandertoorn_twenty_2014]. Individuals would be generally disposed to justify the economic status quo regardless of self- and group economic interest because their support partially depends on their beliefs in system-justifying ideologies [@trump_when_2020]. As a result, what has been suggested is that low-status might perceive the status quo as just as they are more strongly attached to dominant ideologies, while high-status individuals are more prone to justify the system to a greater extent because it favors their interests and dominant position [@jost_theory_2020].

Another factor that has been linked to inequality beliefs is the perception of economic inequality [@kelleyLegitimationInequalityOccupational1993; @osberg_fair_2006;@castillo_perception_2022]. Studies in this matter have argued that perceived economic inequality functions as an anchor for the justifications of inequality because if these inequalities are indeed seen as fair in their generative processes, then they will be considered as just [@wegenerIllusionDistributiveJustice1987; @Wegener1990]. Thus, both the role of social status and economic inequality perceptions have found support in correlational [@Castillo2012a_justice;@schroder_income_2017], as well as in experimental studies that have manipulated inequality perception [@iturra_percepcion_2023;@trump_income_2018]. In addition, normative or value-driven mechanisms can also explain to what extent people justify inequality. For instance, studies suggest that justice ideologies such as individualism or egalitarianism can mediate the influence of individual status on income inequality justification [@Castillo2011]. Likewise, system justification ideologies can strengthen the influence of socioeconomic status on economic inequality preferences [@garcia-sanchez_creencias_2022]. Altogether, these studies have shown that not only socio-structural factors influence what people believe regarding the distribution of economic resources, but their normative stances and perceptions can also determine preferences for economic inequality in diverse domains.

# The justification of educational inequality

@@

The study of inequality justification in the sphere of education is a rather specific and unattended field. Most studies on educational inequality relate performance indicators (such as grades and standardized test results) to stratification-related variables, whereby few refer specifically to if and how individuals tolerate and even justify educational inequalities. This is probably because in principle it seems counter-intuitive to think that someone would be willing to justify differences in such a sensitive area. This paradox is phrased by @confortiLegitimationInequalityAmerican1992 as “while equality is embraced as a value, inequality is accepted as a reality” (p.230).

In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items is diverse, and there is not a single instrument or scale dedicated specifically to this concept. Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables. Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_). Although both questions have been used to measure the justification of educational inequality, they refer to different aspects. The ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive. Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.

Studies have addressed the justification of educational inequality from different conceptual perspectives and methodological strategies. Some perspectives emphasize the justification of differences between specific groups, finding for instance that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps [@valantPoliticsAchievementGaps2016]. Other studies focus on the association between status differences and inequality justification in education, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023]. However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.

The perception of economic inequality is one of the factors associated with the justification of educational inequality. As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023]. For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S. points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality. Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education. Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]

Besides focusing on the relationship between perception and justification of educational inequality, another group of studies has addressed how the justification of inequality is associated with the support of different distributive principles. In an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education. They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school'). In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.

_Justificacion, inequality and meritocracy in the educational domain_

The relationship between performance and outputs brings us to the concepts of merit and meritocracy, key in educational justice research. The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning. Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege. However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020]. This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].

Empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs. Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007]. This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023]. Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes. However, empirical findings also reveal nuances in this relationship. For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023]. This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.

@@

The present research uses longitudinal survey data from 2016 to 2022 to asses changes in inequality justification. As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification in this time frame. Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021]. Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been impacted in this period. Consistently, our first hypothesis is:

>>>>

# The justification of educational inequalities

@@

@@

Research about the justification of educational inequality usually focuses on differences between status groups, based on the rational premise that those who are worst-off would tend to show lower levels of justification [@stoilovaFairnessEducationalOpportunities2023]. For instance, @valantPoliticsAchievementGaps2016 find that Americans are more concerned about—and more supportive of proposals to close—wealth-based achievement gaps than Black-White or Hispanic-White gaps. However, most of the research in this area goes beyond status differences, focusing on how subjective factors such as perceptions and beliefs are related to different levels of justification.

[[ver si esto queda acá o baja]] The perception of economic inequality is one of the factors associated with the justification of educational inequality. As pointed out by @janmaatSubjectiveInequalityReview2013, justifications are usually confused under the labels of “beliefs” and “attitudes,” and their distinction from perceptions is relevant as it helps state that justifying a certain level of inequality requires taking into account which level of (perceived) inequality is first and foremost being recognized [@sonhing_merit_2011;@castillo_multidimensional_2023]. For instance, a recent study by @dayPerceivedIdealInequality2023 on the perception and justification of inequality in university endowments in the U.S. points out that individuals tend to underestimate the magnitude of existing inequality while simultaneously desiring greater equality. Furthermore, through an experiment, they demonstrate that information about inequality in university endowments (i.e., manipulating inequality perception) increases the perception of injustice in education. Results of this study are in line with the consistent evidence on biases related to the underestimation of economic inequality [@castillo_perception_2022;@gimpelson_misperceiving_2018]

In the area of empirical studies, the operationalization of educational inequality justification in terms of questionnaire items has been diverse, and there is not a single instrument or scale dedicated specifically to this concept. Rather, some attitudinal surveys usually include education-related items that are posteriorly used as dependent or independent variables. Some examples are the International Social Survey Programme (ISSP) (_Is it just or unjust – right or wrong – that people with higher incomes can buy better education for their children than people with lower incomes?_), and the European Social Survey (ESS) 2018 (_Compared to other people in [country], I have had a fair chance of achieving the level of education I was seeking_). Although both questions have been used to measure the justification of educational inequality, they refer to different aspects: the ESS can be classified as reflexive, as the respondent is involved as a referent for the question, whereas the ISSP question is non-reflexive. Furthermore, ISSP refers to educational access, while the ESS refers to educational outputs.

_Meritocracy and educational inequalities_

The relationship between performance and outputs brings about to the concepts of merit and meritocracy, key in educational justice research. The concept of meritocracy, coined by the sociologist Michael Young in his satirical work "The Rise of the Meritocracy" [-@young_rise_1962], has a rich history and multifaceted meaning. Young's term was initially used to critique a hypothetical future society where social and economic positions were entirely determined by individuals' merits, talents, and efforts, ostensibly eliminating the influence of inherited privilege. However, over time, the term has evolved to describe a social system where advancement and success are perceived as directly linked to one's abilities and achievements, sidelining inequalities of opportunities [@bell_politics_2020]. This would contribute to the justification of inequality in the educational domain [@sandel_tyranny_2020], as individual ability is often inherited, and educational attainment is influenced by family background [@goldin_meritocracy_2000].

Although there are few studies relating meritocratic beliefs and inequality justification, from a broader perspective it is possible to find related research which incorporates the support of different distributive principles with regard to education. For instance, in an experiment by @igliozziFairShareEffects2024, they establish how the support for different justice principles (equity, equality, and need) is related to support for redistributive policies in domains such as education. They find that a larger system justification  [@jost_psychology_2003] is related to higher support for the principles of equity and equality (over need) in distributing educational outcomes, where equity was measured in terms of access to education related to performance ('Students who place within the highest 50% of testing scores from the previous year can attend the magnet school'). In the same line, @leeFairnessPerceptionsEducational2023 show that performance assessments (in what they call “neoliberal orientations”) predicted people's fair perceptions of educational inequality.

More specific empirical research examining the relationship between meritocratic perceptions and the justification of educational inequality provides valuable insights into the dynamics shaping individuals' attitudes and beliefs. Some studies have explored the extent to which people believe that educational systems are meritocratic and how these perceptions influence their acceptance or rejection of educational disparities, as well as status differences [@McCoy2007]. This belief is particularly influential in educational institutions, where it not only encourages academic success but also legitimizes social and income inequality [@batruch_belief_2023]. Most of the evidence so far suggests that those who believe in a direct link between effort, talent, and educational success are more inclined to rationalize existing disparities in educational outcomes. However, empirical findings also reveal nuances in this relationship. For instance, studies show that when individuals perceive structural barriers, such as economic disparities or discrimination, as hindering the meritocratic ideal, they may be less likely to justify educational inequalities [@dayPerceivedIdealInequality2023]. This suggests that while meritocratic beliefs play a significant role, individuals' awareness of broader societal issues can mediate their willingness to rationalize educational disparities.

@@

The present research aims at identifying changes in the justification of educational inequalities in Chile, with a particular focus in its relationship with meritocratic perceptions. The time frame for this analysis covers between 2016 and 2023, by using longitudinal panel survey data provided by the ELSOC database (Estudio Longitudinal Social de Chile). Although this period comprises the COVID pandemic, the changes in the variables analyzed are not possible to be attributed only to this event in terms of a health crisis, as several economic, social and cultural processes overlapped worldwide. Besides, in the specific case of Chile there were two constitutional processes happening in this very same period after a massive social upheaval. [[redondear-idea]]

As there are measures before and after the COVID pandemic, firstly we explore to what extent it is possible to identify variations in inequality justification within this time frame. Although we are aware that it is not possible to make a causal claim given that this period is characterized by many other circumstances besides COVID, we argue that the increase of risk perceptions in critical situations (such as the pandemic) impacts public concerns about equality and redistribution [@breznau_welfare_2021]. Although the pandemic might have triggered such concerns mostly regarding health, we believe that other policy areas, such as education, could have been affected in this period. Consistently, our first hypothesis is:
